---
title: "NYPD Shooting Analysis"
date: "`r Sys.Date()`"
output: pdf_document
---

## NYPD Shooting Data (Historic)

This document uses the NYPD Historic shooting data located at [NYPD Shooting Data (Historic)](https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD "NYPD Shooting Data (Historic)").

Following is a summary of the raw data set.

```{r load_shooting_data}
# Set the URL and the local file path
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
local_file <- "NYPD_Shooting_Incident_Data.csv"

# Download the file
download.file(url, destfile = local_file)

# Load the data from the downloaded file
shooting_data <- read.csv(local_file, stringsAsFactors = FALSE)

summary(shooting_data)
```

## Cleaning the Data

Next, we will tidy the data to handle missing values, correct data types, and correct/remove anomalies. I am interested in patterns related to the occurrence of incidents by date; are there any apparent seasonal patterns?

```{r clean_shooting_data}
library(dplyr)
library(tidyr)

# Convert date and time columns to datetime objects
shooting_data$OCCUR_DATE <- as.Date(shooting_data$OCCUR_DATE, format = "%m/%d/%Y")
shooting_data$OCCUR_TIME <- hms::as_hms(shooting_data$OCCUR_TIME)

# Remove rows with missing values in critical columns
shooting_data <- shooting_data %>% drop_na(OCCUR_DATE, OCCUR_TIME)

# Remove duplicate rows if any
shooting_data <- shooting_data %>% distinct()

summary(shooting_data)

```

## Visualizing the Data

I am interested in seasonal patterns of shooting incidents, so I will create additional factors to categorize incidents by Year + Month, and Month. Then I plot the seasonal data to look for signs of a seasonal, and the monthly data to visualize the trend irrespective of the year.

```{r visualize_shooting_data}

library(lubridate)
library(ggplot2)

# Extract year-month from the OCCUR_DATE
shooting_data <- shooting_data %>%
  mutate(YearMonth = floor_date(OCCUR_DATE, "month"))

# Count incidents by month
monthly_trend <- shooting_data %>%
  group_by(YearMonth) %>%
  summarize(Incidents = n())

# Plot the annual trend
ggplot(monthly_trend, aes(x = YearMonth, y = Incidents)) +
  geom_line() +
  labs(title = "Trend of Shooting Incidents Over Time", x = "Year-Month", y = "Number of Incidents")

# Extract the month from the OCCUR_DATE
shooting_data <- shooting_data %>%
  mutate(Month = month(OCCUR_DATE, label = TRUE))

# Aggregate incidents by month
monthly_distribution <- shooting_data %>%
  group_by(Month) %>%
  summarize(Incidents = n())

# Plot the monthly distribution
ggplot(monthly_distribution, aes(x = Month, y = Incidents, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Seasonal Pattern of Shooting Incidents by Month", x = "Month", y = "Number of Incidents") +
  theme_minimal()

```

## Analyzing the Data

From the visualization it is seems that shooting incidents peak in the summer and are at their lowest in the winter months. I can also see a trend of shooting incidents decreasing from 2009 until 2020, when they quickly spike.

```{r seasonal_visualization}

# Create a time series object with monthly data
ts_data <- ts(monthly_trend$Incidents, frequency = 12)

# Perform STL decomposition
stl_decomposition <- stl(ts_data, s.window = "periodic")

# Plot the components
plot(stl_decomposition)

```

From this decomposition, we can more specifically look at the overall trend by year.

```{r seasonal_analysis}

# Extract components
seasonal_component <- stl_decomposition$time.series[, "seasonal"]
trend_component <- stl_decomposition$time.series[, "trend"]
residual_component <- stl_decomposition$time.series[, "remainder"]

# Generate the corresponding dates for the time series
start_date <- min(monthly_trend$YearMonth)
end_date <- max(monthly_trend$YearMonth)
time_points <- seq.Date(from = start_date, to = end_date, by = "month")

# Create indices so we can label the x-axis by year
time_index <- seq_along(trend_component)

# Plot the trend component
plot(time_index, trend_component, type = "l", col = "red", lwd = 2,
     main = "Overall Trend", ylab = "Trend", xlab = "Year", xaxt = "n")

# Add custom X-axis labels at appropriate intervals
axis(1, at = time_index[seq(1, length(time_index), by = 12)], 
     labels = format(time_points[seq(1, length(time_points), by = 12)], "%Y"), las = 2)

```

## Modeling the Data

After separating the data into seasonal and trend components, we can produce models of each. For the seasonal component, we can use a Fourier transform to model the harmonic pattern the data shows over the years:

```{r modeling_seasonal}

# Given monthly data with annual seasonality (period = 12)
period <- 12

# Create sine and cosine terms for the harmonic model
harmonics <- cbind(sin(2 * pi * time_index / period), cos(2 * pi * time_index / period))

# Fit the harmonic regression model
seasonal_model <- lm(seasonal_component ~ harmonics)
summary(seasonal_model)

plot(time_index, seasonal_component, type = "l", col = "red", lwd = 2,
     main = "Seasonal Model", ylab = "Trend", xlab = "Year", xaxt = "n") 

# Add custom X-axis labels at appropriate intervals
axis(1, at = time_index[seq(1, length(time_index), by = 12)], 
     labels = format(time_points[seq(1, length(time_points), by = 12)], "%Y"), las = 2)

```

The overall trend is more complex and likely the product of many factors not available in the source data. For instance, the Covid pandemic seems to have been a major factor in the [increase of shootings during 2020 in the US](https://everytownresearch.org/report/gun-violence-and-covid-19-in-2020-a-year-of-colliding-crises/).

While a more complex model might reflect this increase, looking at data beyond 2020 indicates that the overall trend of decreased shootings seems to continue after this spike. That in mind, I will instead use a simpler linear model, knowing to expect a large variance around 2020.

```{r modeling_trend}

# Fit a simple linear regression model
trend_model_linear <- lm(trend_component ~ time_index)
summary(trend_model_linear)

# Plot the trend with the fitted linear model
plot(time_index, trend_component, type = "l", col = "red", lwd = 2,
     main = "Trend Component with Fitted Linear Model", ylab = "Trend", xlab = "Year", xaxt="n")
abline(trend_model_linear, col = "blue", lwd = 2)  # Add the linear fit

# Add custom X-axis labels at appropriate intervals
axis(1, at = time_index[seq(1, length(time_index), by = 12)], 
     labels = format(time_points[seq(1, length(time_points), by = 12)], "%Y"), las = 2)

```

I can now combine the seasonal model with the trend model to produce a combined model:

```{r combined_model}

# Predict the trend component using the linear model
trend_fitted <- predict(trend_model_linear)

# Predict the seasonal component using the harmonic model
seasonal_fitted <- predict(seasonal_model)

# Combine the trend and seasonal predictions
combined_fitted <- trend_fitted + seasonal_fitted

# Plot the actual data
plot(time_index, ts_data, type = "l", col = "black", lwd = 2,
     main = "Actual Data vs. Combined Trend and Seasonal Model", ylab = "Shootings", xlab = "Year", xaxt = "n")

# Add the combined model
lines(time_index, combined_fitted, col = "blue", lwd = 2)

# Add the trend component
lines(time_index, trend_fitted, col = "red", lwd = 2, lty = 2)

# Add X-axis labels
axis(1, at = time_index[seq(1, length(time_index), by = 12)], 
     labels = format(time_points[seq(1, length(time_points), by = 12)], "%Y"), las = 2)

# Add a legend to differentiate the lines
legend("top", legend = c("Actual Data", "Combined Model", "Trend Component"),
       col = c("black", "blue", "red"), lty = c(1, 1, 2), lwd = 2)

# Calculate residuals
residuals <- ts_data - combined_fitted

# Plot a histogram of residuals to assess their distribution
hist(residuals, breaks = 20, col = "gray", main = "Histogram of Residuals", xlab = "Residuals")

```

## Bias Identification

Some unknowns that may affect this model are the thoroughness of the underlying data, say if standards for reporting have changed over time or varied within certain precincts.

The model does not consider other factors that may be important, such as economic events, political activity, etc. I chose a simple linear regression model for the overall trend to avoid overfitting the model around the occurrence of the pandemic, and including this data suggests that had the pandemic never occurred, we might expect an even steeper reduction in shootings over this period.